Deep mutational scanning of hemagglutinin helps identify evolutionarily successful human H3N2 influenza viruses

### 2018-01-30 paper/manuscript.tex
added some Mike comments

% Please add here a significance statement to explain the relevance of your work
% 120 words max
\significancestatement{ 
A key goal in the study of influenza virus evolution is to forecast which viral strains will persist and which ones will die out. 
Here we experimentally measure the effects of all amino-acid mutations to the hemagglutinin protein from a human H3N2 influenza strain on viral growth in cell culture.
We show that these measurements have utility for distinguishing among viral strains that do and do not succeed in nature.
Overall, our work suggest that new high-throughput experimental approaches may be useful for evolutionary forecasting.
}

@@

We next quantified the reproducibility of our deep mutational scanning measurements across biological and technical replicates. 
We first used the deep sequencing data for each replicate to estimate the preference of each site in HA for all 20 amino acids using the method described in~\cite{bloom2015software}.
Because there are 566 residues in HA, there are $566 \times 20 = 11,320$ estimated amino-acid preferences, corresponding to $566 \times 19 = 10,754$ distinct measurements (the 20 preferences at each site are constrained to sum to one~\cite{bloom2015software}).
The correlations of the amino-acid preferences between pairs of replicates are shown in Figure~\ref{fig:dms_overview}D.
The biological replicates are fairly well-correlated, with Pearson's $R$ ranging from 0.69 to 0.78. 
Replicate 1 exhibited the weakest correlation with the other replicates; this replicate also showed the weakest selection against stop and nonsynonymous mutations (Figure~\ref{fig:dms_overview}C), perhaps indicating more experimental noise.
The two technical replicates 3-1 and 3-2 were only slightly more correlated than pairs of biological replicates, suggesting that bottlenecking of library diversity after the reverse-genetics step contributes most of the experimental noise.

@@

Our experiments were performed on the Perth/2009 HA, but we scored mutations on a phylogenetic tree that extends from 1968 to 2017.
Because the effects of mutations can vary with genetic background~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative}, it is possible that the effects of mutations that we measured in the Perth/2009 HA might have shifted somewhat in other related HAs.
To explore this question, we scored the complete HA sequence $\mathbf{s}$ of every node in the phylogenetic tree by quantifying its average per-site sequence preference as 
\begin{equation}
F\left(\mathbf{s}\right) = \frac{1}{L}\displaystyle\sum_{r=1}^L \ln \pi_{r, s_r},
\end{equation}
where $\pi_{r, s_r}$ is the preference measured in our experiments (e.g., Figure~\ref{fig:logoplot}) for the amino acid $s_r$ at site $r$ in HA sequence $\mathbf{s}$, and $L$ is the length of the sequence.
Figure~\ref{fig:trunkvssidebranch}C shows that the sequences of trunk nodes tend to have higher preference than those of side branch nodes across the entire timespan, consistent with the finding that trunk mutations are generally more favorable than side branch mutations. 
However, the average per-site sequence preference increases as the nodes approach the Perth/2009 strain.
The Perth/2009 itself has among the highest per-site sequence preference of the entire tree, despite falling on a side branch.
We do not think that this result means that Perth/2009 is actually higher fitness than the other strains.
Rather, if there are modest shifts in the effects of mutations as HA evolves, then sequences more distant from Perth/2009 could be scored as having lower preference simply because these shifts start to degrade the accuracy of our measurements.
But importantly, our experiments consistently show the trunk to be more favorable than the side branches regardless of whether we look at sequences that precede Perth/2009 (most of Figure~\ref{fig:trunkvssidebranch}C) or sequences that occur after Perth/2009 has diverged from the trunk (Figure~\ref{suppfig:future}). 
Therefore, our experimental measurements consistently reveal the selective advantage of the trunk over more than a half-century of viral evolution.

@@

Mutations that occur on the trunk are scored as significantly more favorable in our experiments than side-branch mutations at both epitope and non-epitope sites (Figure~\ref{fig:sequence_preference}A).
Therefore, our experiments can distinguish evolutionarily favorable mutations both at sites that are predominantly under functional constraint and at sites that are also under immune pressure.
The fact that our experiments can discriminate between the trunk and side branches at epitope sites despite only assaying for viral growth underscores the fact that most epitope sites are still important for HA function~\cite{nakajima2003restriction,das2013defining,koel2013substitutions}.

@@

To more rigorously quantify shifts in amino-acid preferences after correcting for experimental noise, we used the statistical approach in~\cite{doud2015site,haddox2017mapping}.
Figure~\ref{fig:distance_distribution}C shows the distribution of shifts in amino-acid preferences between H3 and H1 HAs after correcting for experimental noise.
Although some sites have small shifts near zero, many sites have large shifts.
These shifts between H3 and H1 are much larger than expected from the null distribution that would be observed purely from experimental noise.
They are also much larger than the shifts previously observed between two HIV Envs with 86\% amino-acid identity~\cite{haddox2017mapping}.
However, the shifts are still smaller than those observed when comparing HA to the non-homologous HIV Env protein. 
Therefore, there are very substantial shifts in mutational effects between highly diverged HA homologs, although the effects of mutations remain more similar than for non-homologous proteins.

\begin{SCfigure*}[\sidecaptionrelwidth][t]
\centering
\includegraphics[width=12cm]{figs/distance_distribution/distance_distribution.pdf}
\caption{\label{fig:distance_distribution}
{\bf There are substantial differences in the effects of mutations between H1 and H3 HAs.}
(A) Phylogenetic tree of HA subtypes, with the WSN/1933 H1 and Perth/2009 H3 HAs labeled. 
These HAs have 42\% amino-acid identity.
(B) All pairwise correlations of the amino-acid preferences measured in the three individual deep mutational scanning replicates in the current study  and the three replicates in prior deep mutational scanning of an H1 HA~\cite{doud2016accurate}.
Comparisons between H3 replicates are in purple, those between H1 replicates are in brown, and those across H1 and H3 replicates are in gray. 
$R$ indicates the Pearson correlation coefficient.
(C) We calculated the shift in amino-acid preferences at each site between H3 and H1 HAs using the method in \cite{haddox2017mapping}, and plotted the distribution of shifts for all sites.
The shifts between H3 and H1 (yellow) are much larger than the null distribution (blue) expected if all differences are due to experimental noise.
The shifts are also much larger than those previously observed between two variants of HIV envelope protein (Env) that share 86\% amino-acid identity (pink).
However, the shifts between H3 and H1 are still less than the differences between either HA and HIV Env (green).
}
\end{SCfigure*}

@@

Our work did not characterize the antigenic effects of mutations, which also play an important role in determining strain success in nature~\cite{koel2013substitutions,neher2016prediction}.
Indeed, some of our results (\textit{e.g.}, Figure~\ref{fig:sequence_preference}) suggest as much by showing that the experiments better capture selection at non-epitope than epitope sites.
However, our basic selection and deep-sequencing approach can be harnessed to completely map how mutations affect antibody recognition~\cite{doud2017complete,doud2017quantifying}.
But so far, experiments using this approach have either failed to examine antibodies or sera that are relevant to driving the evolution of H3N2 influenza~\cite{doud2017complete,doud2017quantifying}, or have used relevant sera but examined a non-comprehensive set of mutations~\cite{li2016selection}.
Future experiments that completely map how HA mutations affect recognition by human sera seem likely to be especially fruitful for informing viral forecasting.

>>>>

% Please add here a significance statement to explain the relevance of your work
% 120 words max
\significancestatement{ 
A key goal in the study of influenza virus evolution is to forecast which viral strains will persist and which ones will die out. 
Here we experimentally measure the effects of all amino-acid mutations to the hemagglutinin protein from a human H3N2 influenza strain on viral growth in cell culture.
We show that these measurements have utility for distinguishing among viral strains that do and do not succeed in nature.
Overall, our work suggests that new high-throughput experimental approaches may be useful for evolutionary forecasting.
}

@@

We next quantified the reproducibility of our deep mutational scanning measurements across biological and technical replicates. 
We first used the deep sequencing data for each replicate to estimate the preference of each site in HA for all 20 amino acids using the method described in~\cite{bloom2015software}.
Because there are 566 residues in HA, there are $566 \times 20 = 11,320$ estimated amino-acid preferences, corresponding to $566 \times 19 = 10,754$ distinct measurements (the 20 preferences at each site are constrained to sum to one~\cite{bloom2015software}).
The correlations of the amino-acid preferences between pairs of replicates are shown in Figure~\ref{fig:dms_overview}D.
The biological replicates are fairly well-correlated, with Pearson's $R$ ranging from 0.69 to 0.78. 
Replicate 1 exhibited the weakest correlation with the other replicates; this replicate also showed the weakest selection against stop and nonsynonymous mutations (Figure~\ref{fig:dms_overview}C), perhaps indicating more experimental noise.
The two technical replicates 3-1 and 3-2 were only slightly more correlated than pairs of biological replicates, suggesting that bottlenecking of library diversity during viral passage contributes most of the experimental noise.

@@

Our experiments were performed on the Perth/2009 HA, but we scored mutations on a phylogenetic tree that extends from 1968 to 2017.
Because the effects of mutations can vary with genetic background~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative}, it is possible that the effects of mutations that we measured in the Perth/2009 HA might have shifted somewhat in other related HAs.
To explore this question, we scored the complete HA sequence $\mathbf{s}$ of every node in the phylogenetic tree by quantifying its average per-site sequence preference as 
\begin{equation}
F\left(\mathbf{s}\right) = \frac{1}{L}\displaystyle\sum_{r=1}^L \ln \pi_{r, s_r},
\end{equation}
where $\pi_{r, s_r}$ is the preference measured in our experiments (e.g., Figure~\ref{fig:logoplot}) for the amino acid $s_r$ at site $r$ in HA sequence $\mathbf{s}$, and $L$ is the length of the sequence.
Figure~\ref{fig:trunkvssidebranch}C shows that the sequences of trunk nodes tend to have higher preference than those of side branch nodes across the entire timespan, consistent with the finding that trunk mutations are generally more favorable than side branch mutations. 
However, the average per-site sequence preference increases as the nodes approach the Perth/2009 strain.
The Perth/2009 itself has among the highest per-site sequence preference of the entire tree, despite falling on a side branch.
We do not think that this result means that Perth/2009 is actually higher fitness than the other strains.
Rather, if there are modest shifts in the effects of mutations as HA evolves, then sequences more distant from Perth/2009 could be scored as having lower preference simply because these shifts start to degrade the applicability of our measurements.
But importantly, our experiments consistently show the trunk to be more favorable than the side branches regardless of whether we look at sequences that precede Perth/2009 (most of Figure~\ref{fig:trunkvssidebranch}C) or sequences that occur after Perth/2009 has diverged from the trunk (Figure~\ref{suppfig:future}). 
Therefore, our experimental measurements consistently reveal the selective advantage of the trunk over more than a half-century of viral evolution.

@@

Mutations that occur on the trunk are scored as significantly more favorable in our experiments than side-branch mutations at both epitope and non-epitope sites (Figure~\ref{fig:sequence_preference}A).
Therefore, our experiments can distinguish evolutionarily favorable mutations both at sites that are predominantly under functional constraint and at sites that are also under immune pressure.
The fact that our experiments can discriminate between the trunk and side branches at epitope sites despite only assaying for viral growth underscores the fact that mutations at epitope sites still have significant non-antigenic functional consequences~\cite{nakajima2003restriction,das2013defining,koel2013substitutions}.

@@

To more rigorously quantify shifts in amino-acid preferences after correcting for experimental noise, we used the statistical approach in~\cite{doud2015site,haddox2017mapping}.
Figure~\ref{fig:distance_distribution}C shows the distribution of shifts in amino-acid preferences between H3 and H1 HAs after correcting for experimental noise.
Although some sites have small shifts near zero, many sites have large shifts.
These shifts between H3 and H1 are much larger than expected from the null distribution that would be observed purely from experimental noise.
They are also much larger than the shifts previously observed between two HIV Envs with 86\% amino-acid identity~\cite{haddox2017mapping}.
However, the typical shift between H3 and H1 is still smaller than that observed when comparing HA to the non-homologous HIV Env protein. 
Therefore, there are very substantial shifts in mutational effects between highly diverged HA homologs, although the effects of mutations remain more similar than for non-homologous proteins.

\begin{SCfigure*}[\sidecaptionrelwidth][t]
\centering
\includegraphics[width=12cm]{figs/distance_distribution/distance_distribution.pdf}
\caption{\label{fig:distance_distribution}
{\bf There are substantial differences in the effects of mutations between H1 and H3 HAs.}
(A) Phylogenetic tree of HA subtypes, with the WSN/1933 H1 and Perth/2009 H3 HAs labeled. 
These HAs have 42\% amino-acid identity.
(B) All pairwise correlations of the amino-acid preferences measured in the three individual deep mutational scanning replicates in the current study  and the three replicates in prior deep mutational scanning of an H1 HA~\cite{doud2016accurate}.
Comparisons between H3 replicates are in purple, those between H1 replicates are in brown, and those across H1 and H3 replicates are in gray. 
$R$ indicates the Pearson correlation coefficient.
(C) We calculated the shift in amino-acid preferences at each site between H3 and H1 HAs using the method in \cite{haddox2017mapping}, and plotted the distribution of shifts for all sites.
The shifts between H3 and H1 (yellow) are much larger than the null distribution (blue) expected if all differences are due to experimental noise.
The shifts are also much larger than those previously observed between two variants of HIV envelope protein (Env) that share 86\% amino-acid identity (pink).
However, the shifts between H3 and H1 are usually less than the differences between either HA and HIV Env (green).
}
\end{SCfigure*}

@@

Our work did not characterize the antigenic effects of mutations, which also play an important role in determining strain success in nature~\cite{koel2013substitutions,neher2016prediction}.
%Indeed, some of our results (\textit{e.g.}, Figure~\ref{fig:sequence_preference}) suggest as much by showing that the experiments better capture selection at non-epitope than epitope sites.
However, our basic selection and deep-sequencing approach can be harnessed to completely map how mutations affect antibody recognition~\cite{doud2017complete,doud2017quantifying}.
But so far, experiments using this approach have not examined antibodies or sera that are relevant to driving the evolution of H3N2 influenza~\cite{doud2017complete,doud2017quantifying}, or have used relevant sera but examined a non-comprehensive set of mutations~\cite{li2016selection}.
Future experiments that completely map how HA mutations affect recognition by human sera seem likely to be especially fruitful for informing viral forecasting.
